### 2021-09-14 paper/localpsf.tex
fixed a lot of notational collisions

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain in dimension $d=1$, $2$, or $3$. We consider integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ of the form

@@

If $\Aop$ were perfectly translation invariant (i.e., if equality held in \eqref{eq:local_translation_invariance} for all $x$, $y$, $h$) then it is straightforward to show that $\Aop$ would be the convolution operator $u \mapsto \left(\convkernel_x \ast u\right)^*$, where the convolution kernel is given by
\begin{equation}
	\label{eq:convolution_kernel}
	\convkernel_x(y) := \impulseresponse_x(y+x).
\end{equation}
The convolution kernel $\convkernel_x$ is the result of translating the impulse response $\impulseresponse_x$ to re-center it at zero instead of $x$. Locally translation invariant operators act like convolution operators locally, but the convolution kernel, $\convkernel_x$, varies as $x$ changes. We therefore approximate $\Aop$ by a spatially varying weighted sum of convolution operators, $\AopPc \approx \Aop$, of the form
\begin{equation}
\label{eq:product_convolution}
	\left(\AopPc~u\right)(v) := \left( v,~ \sum_{i=1}^r \convkernel_i \ast \left(w_i \cdot u\right) \right)_{L^2(\Omega)},
\end{equation}
where $f \cdot g$ denotes pointwise multiplication, $f \ast g$ denotes convolution, and $\left(f, g\right)_{L^2(\Omega)}$ denotes the $L^2$ inner product on $\Omega$, for functions $f,g \in L^2(\Omega)$. Approximations of the form \eqref{eq:product_convolution} are known as \emph{product-convolution} approximations, because the action of each term in the sum consists of a pointwise product, followed by a convolution. To avoid nested subscripts, we write 
\begin{equation*}
	\convkernel_i := \convkernel_{x_i} \quad \text{and} \quad \impulseresponse_i := \impulseresponse_{x_i},
\end{equation*}
with $\convkernel_{x_i}$ defined in \eqref{eq:convolution_kernel} and $\impulseresponse_{x_i}$ defined in \eqref{eq:impulse_response_defn}, to denote local convolution kernels and impulse responses corresponding to a collection of points $\{x_i\}_{i=1}^r \subset \Omega$ scattered throughout the domain. The functions $w_i$ are spatially varying weighting functions that are used to interpolate the convolution kernels $\convkernel_i$. Our operator approximation method is defined by how we compute the convolution kernels, $\convkernel_i$, how we choose the points, $x_i$, and what weighting functions, $w_i$, that we use. In Section \ref{sec:overview_intro} we summarize the answers to these questions; we will provide detailed answers to these questions in Section \ref{sec:method}.

@@

In order to choose the points $x_i$, we need to estimate the support of $\impulseresponse_i$ \emph{before} we compute it. The supports of the functions $\impulseresponse_x$ are estimated a-priori, for all $x \in \Omega$ simultaneously, via a procedure that involves applying $\Aop^T$ to a small number of specially chosen functions (three functions if $d=1$, six if $d=2$, ten if $d=3$), then post processing the results (Section \ref{eq:mean_and_covariance_estimation}). This procedure is the only part of the paper that requires non-negativity of the integral kernel.

@@

Once the $\convkernel_i$ and $w_i$ are computed, we convert $\AopPc$ to hierarchical matrix (H-matrix) format. Let $\Aker_\text{pc}$ denote the integral kernel associated with $\AopPc$. While $\Aker(y,x)$ is not easily computable, $\Aker_\text{pc}(y,x)$ is given by the formula
\begin{equation}
	\label{eq:kernel_entries}
	\AkerPc(y,x) = \sum_{i=1}^r \convkernel_i(y-x) w_i(x).
\end{equation}
This follows from~\eqref{eq:product_convolution} and the fact that the convolution operator $u \mapsto \convkernel_i \ast u$ has $(y,x)$ kernel entry given by $\convkernel(y-x)$.
Formula~\eqref{eq:kernel_entries} allows us to construct a H-matrix representation of $\AopPc$ using the conventional adaptive cross H-matrix construction method, in which one forms low-rank approximations of blocks of the matrix by sampling rows and columns of those blocks \cite{HACA}. Once in H-matrix format, fast H-matrix arithmetic is used to invert $\AopPc$, or perform other useful matrix operations.

@@

		\For{$b=1,2,\dots,\text{max\_batches}$}{
			Choose a batch of sample points, $\pointbatch_b$, using Algorithm \ref{alg:point_choice}

			Compute $\combresponse_b$ by applying $\Aop$ to the dirac comb for $\pointbatch_b$ (Section \ref{sec:get_impulse_response})

@@

The ellipsoid $E_x$ is not immediately available. To use $E_x$, we must first compute the scalar $\spatialvol(x)$, the length-$d$ vector $\spatialmean(x)$, and the $d \times d$ matrix $\spatialcov(x)$ (recall $d \in \{1,2,3\}$ is the spatial dimension of $\Omega$). The direct approach to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$. But this direct approach is computationally infeasible because and it requires one operator action of $\Aop$ per point $x$ that we wish to evaluate $E_x$ at, and we will need to evaluate $E_x$ at a large number of points. Fortunately, it is possible to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$, \emph{for all points $x$ simultaneously}, by applying $\Aop^T$ to one constant function, $d$ linear functions, and $d(d+1)/2$ quadratic functions. We present this efficient method of computing $\spatialvol$, $\spatialmean$, and $\spatialcov$ in Theorem \ref{thm:vol_mean_cov} and summarize the method in Algorithm \ref{alg:varhpi_mean_cov}.

\begin{theorem}[Impulse response moments]
	\label{thm:vol_mean_cov}
	Let $\spatialvol:\Omega \rightarrow \mathbb{R}$, $\spatialmean:\Omega \rightarrow \mathbb{R}^d$, and $\spatialcov:\Omega \rightarrow \mathbb{R}^{d \times d}$, be the following functions:
	\begin{equation*}
		\spatialvol(x) := \int_{\Omega} \impulseresponse_x(y) dy, \qquad
		\spatialmean(x) := \mathbb{E}(\widehat{\impulseresponse}_x), \qquad
		\spatialcov(x) := \Var(\widehat{\impulseresponse}_x),
	\end{equation*}
	where $\widehat{\impulseresponse}_x := \impulseresponse_x / \spatialvol(x)$ denotes the normalized version of $\impulseresponse_x$, $\mathbb{E}(\widehat{\impulseresponse}_x)$ denotes the mean of $\widehat{\impulseresponse}_x$, and $\Var(\widehat{\impulseresponse}_x)$ denotes the variance of $\widehat{\impulseresponse}_x$.	Let  $x^i$ denote the $i^\text{th}$ component of $x$, i.e., $x = \left(x^1, x^2, \dots, x^d\right)$. Let $C$, $\{L^i\}_{i=1}^d$, and ${\{Q^{ij}\}_{i=1}^d}_{j=1}^d$ be the following constant, linear, and quadratic functions:
	\begin{equation*}
		C(x) := 1, \qquad
		L^i(x) := x^i, \qquad
		Q^{ij}(x) = x^i x^j.
	\end{equation*}
	Let $f/g$ denote the pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ denote the pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$. The following results hold:
	\begin{subequations}
		\label{eq:vol_mean_var}
		\begin{align}
			\spatialvol =& \left(\Aop^T C\right)^* \\
			\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
			\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
		\end{align}
	\end{subequations}
	 for $i=1,\dots, d$, $j=1,\dots,d$.
\end{theorem}

@@

		\tcp{Compute mean $\spatialmean$}
		\For{$i=1,2,\dots,d$}{
			Form linear function $L^i(x) = x^i$

			Compute $\spatialmean^i = \left(\Aop^T L^i\right)^* / \spatialvol$
		}
		\tcp{Compute covariance $\spatialcov$}
		\For{$i=1,2,\dots,d$}{
			\For{$j=1,\dots,i$}{
				Form quadratic function $Q^{ij}(x) = x^i x^j$

@@

We use a greedy ellipsoid packing algorithm to choose sets of sample points, $\pointbatch_b$, such that there is no overlap between the ellipsoids, $E_{x_i}$, associated with the sample points, $x_i$, within a batch. These sample point batches will be used in Section \ref{sec:get_impulse_response} to compute many impulse responses of $\Aop$ using only a small number of operator actions of $\Aop$. The support of $\impulseresponse_x$ is (approximately) contained in the ellipsoid $E_x$, so by applying $\Aop$ to the Dirac comb associated with all sample points in a batch, we will recover the impulse responses for all sample points in that batch.

We start with a finite set of candidate points $P$. To build a batch of sample points, $\pointbatch_b$, first we initialize $\pointbatch_b$ as an empty set. Then we select the point $p \in P$ that is the farthest away from all points in previous sample point batches. That is, $p$ is a maximizer of the following optimization problem:
\begin{equation*}
\max_{p \in P} \min_{x \in \pointbatch_1 \cup \dots \cup \pointbatch_{b-1}} \|p - x\|.
\end{equation*}
Candidate points for the first batch are chosen arbitrarily from $P$.
Once $p$ is selected, we remove $p$ from $P$. If $p$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_p \cap E_q = \{\}$ for all $q \in \pointbatch_b$, then we add $p$ to $\pointbatch_b$. Otherwise we discard $p$. This process repeats until there are no more points in $P$.  This is detailed in Algorithm \ref{alg:point_choice}. We determine whether $E_p \cap E_q = \{\}$ using the fast ellipsoid intersection test described in Appendix \ref{sec:fast_ellipsoid_intersection_test}.

We repeat the process to construct several batches of points $\pointbatch_1, \pointbatch_2, \dots$, until the number of batches exceeds a desired threshold. In our implementation, for each batch the set of candidate points $P$ is initialized as the set of all Lagrange nodes for the finite element basis functions used to discretize the problem, minus all points in previously chosen batches.

@@

	\Input{Finite set of candidate points $P\subset \Omega$, \\spatialmean $\spatialmean(x)$ and covariance $\spatialcov(x)$, \\domain boundary information $\partial \Omega$, \\previous sample point batches $\pointbatch_1, \dots, \pointbatch_{b-1}$}
	\Output{Batch of new sample points $\pointbatch_b$}

	Initialize empty new batch of sample points, $\pointbatch_b = \{\}$

	\While{$P$ is not empty}{
		Determine the point $p \in P$ that is farthest from all points in the previous sample point batches $\pointbatch_1,\dots,\pointbatch_{b-1}$

@@

		\If{$E_p \cap E_q \neq \{\}$ for all $q \in \pointbatch_b$ and $E_p \cap \partial \Omega = \{\}$}{
			\tcp{$E_p$ and $E_q$ are the ellipsoids defined in \eqref{eq:support_ellipsoid}}

			Add $p$ to $\pointbatch_b$

@@

    \SetKwFunction{FMain}{point\_is\_acceptable}
	\caption{Choosing one batch of sample points, $\pointbatch_b$}
	\label{alg:point_choice}
\end{algorithm2e}

@@

We compute convolution kernels $\convkernel_{x_i}$ in batches by applying $\Aop$ to a small number Dirac combs. Each application of $\Aop$ to a Dirac comb yields one batch of impulse responses, $\impulseresponse_{x_i}$, which yields one batch of convolution kernels, $\convkernel_{x_i}$. The Dirac comb, $\dcomb_b$, associated with a batch of sample points, $\pointbatch_b$, is the sum of Dirac distributions (point sources) centered at the points $x_i \in \pointbatch_b$. That is,
\begin{equation*}
	\dcomb_b := \sum_{x_i \in \pointbatch_b} \delta_{x_i}.
\end{equation*}
For each batch $\pointbatch_b$, we compute the action of $\Aop$ on the associated Dirac comb:
\begin{equation}
	\label{eq:dirac_comb_H_action}
	\combresponse_b := \left(\Aop \dcomb_b^*\right)^*.
\end{equation}
By linearity, the resulting function $\combresponse_b$ may be written as
\begin{equation}
	\label{eq:phi_b}
	\combresponse_b = \left(\Aop \left(\sum_{x_i \in \pointbatch_b} \delta_{x_i}\right)^*\right)^* = \sum_{x_i \in \pointbatch_b} \left(\Aop \delta_{x_i}^*\right)^* = \sum_{x_i \in \pointbatch_b} \impulseresponse_{x_i}.
\end{equation}
Since the points $x_i$ are chosen so that the ellipsoid $E_{x_i}$ that (approximately) supports $\impulseresponse_i$, and the ellipsoid $E_{x_j}$ that (approximately) supports $\impulseresponse_j$ do not overlap when $i \neq j$, from \eqref{eq:convolution_kernel} and \eqref{eq:phi_b} we have (approximately)
\begin{equation}
\label{eq:varphi_eval}
	\convkernel_{x_i}(z) = \impulseresponse_i(z+x_i) = 
	\begin{cases}
		\combresponse_b(z+x_i), & z+x_i \in E_{x_i} \\
		0, & \text{otherwise}
	\end{cases}
\end{equation}
for all $x_i \in \pointbatch_b$. By performing one matrix-vector product, \eqref{eq:dirac_comb_H_action}, we recover the convolution kernels $\convkernel_{x_i}$ associated with every point $x_i \in \pointbatch_b$.

@@

To convert $\AopPc$ to H-matrix format (Section \ref{sec:H_matrix_conversion}), we will need to evaluate $\convkernel_{x_i}(z)$ for a large number of points $z$. Per \eqref{eq:varphi_eval}, evaluating $\convkernel_{x_i}(z)$ requires evaluating $\combresponse_b(z+x_i)$. In computations, $\combresponse_b$ will be a finite element function formed on an irregular mesh, and $z+x_i$ will typically not be a gridpoint of the mesh. Evaluating $\combresponse_b(z+x_i)$ therefore requires finding which mesh cell that $z+x_i$ is in, then evaluating several polynomial functions to evaluate the finite element function on that cell. Although this procedure is asymptotically scalable, it is slow in practice \cite{FEMEVALSLOW}.

To speed up evaluation of $\convkernel_{x_i}$, we interpolate $\convkernel_{x_i}$ onto a regular rectilinear grid on a coordinate axis aligned box that just barely contains the ellipsoid $E_{x_i}$. We transfer the function $\convkernel_{x_i}$ from the irregular mesh to the regular grid by evaluating $\convkernel_{x_i}$ at all gridpoints in the regular grid, using the slow process described in the previous paragraph. This is an upfront cost which is done ``offline.'' After $\convkernel_{x_i}$ has been transferred to the regular grid, we use fast regular grid interpolation to perform all subsequent ``online'' evaluations of $\convkernel_{x_i}$. This yields a substantial overall speedup because the number of finite element function evaluations required to transfer the functions $\convkernel_{x_i}$ to the regular grid is $O(N)$, while the number of kernel evaluations required to convert $\AopPc$ to H-matrix format is $O(k N \log N)$. Here $N$ is the number of finite element degrees of freedom, and $k$ is the H-matrix rank. Furthermore, transferring $\convkernel_{x_i}$ to a regular grid requires evaluating $\convkernel_{x_i}$ at a predetermined and highly structured collection of points. This structure allows for faster determination of which mesh cells that points are in, and allows one to evaluate the finite element function at all regular gridpoints in parallel. In contrast, constructing the H-matrix requires evaluating $\convkernel_{x_i}$ at an unpredictable sequence of scattered of points, which is an expensive task for finite element function evaluation, but a cheap task for regular grid interpolation. 
Having $\convkernel_{x_i}$ stored on a regular grid also simplifies the implementation of the method that we use to deal with boundary artifacts, which will be described in Section \ref{sec:boundary_extension}.

@@

We choose the $i^\text{th}$ weighting function, $w_i(x)$, to be the solution to the following optimization problem:
\begin{equation}
	\label{eq:wi_optimization_problem}
	\begin{aligned}
		\min_{w_i \in H^2_N(\Omega)} &\quad \frac{1}{2}\norm{\Delta w_i}_{L^2(\Omega)}^2 \\
		\text{such that} &\quad w_i(x_j) = \delta_{ij}, \quad j=1,\dots,r.
	\end{aligned}
\end{equation}
Here $\Delta$ is the Laplacian on $\Omega$, $\delta_{ij}$, is the Kronecker delta, which takes value one if $i=j$ and zero if $i \neq j$, and $H^2_N(\Omega)$ is the space of all functions in $H^2(\Omega)$ with Neumann zero boundary conditions. Using these weighting functions yields the ``flattest possible'' interpolation of the convolution kernels $\convkernel_i$, in a least-squares sense (Theorem \ref{thm:smoothest_interpolant}). Also, these weighting functions sum to one pointwise (Proposition \ref{thm:wi_sum_to_one}).

>>>>

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain in dimension $\gdim=1$, $2$, or $3$. We consider integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ of the form

@@

If $\Aop$ were perfectly translation invariant (i.e., if equality held in \eqref{eq:local_translation_invariance} for all $x$, $y$, $h$) then it is straightforward to show that $\Aop$ would be the convolution operator $u \mapsto \left(\convkernel_x \ast u\right)^*$, where the convolution kernel is given by
\begin{equation}
	\label{eq:convolution_kernel}
	\convkernel_x(y) := \impulseresponse_x(y+x).
\end{equation}
The convolution kernel $\convkernel_x$ is the result of translating the impulse response $\impulseresponse_x$ to re-center it at zero instead of $x$. Locally translation invariant operators act like convolution operators locally, but the convolution kernel, $\convkernel_x$, varies as $x$ changes. We therefore approximate $\Aop$ by a spatially varying weighted sum of convolution operators, $\AopPc \approx \Aop$, of the form
\begin{equation}
\label{eq:product_convolution}
	\left(\AopPc~u\right)(v) := \left( v,~ \sum_{i=1}^\convrank \convkernel_i \ast \left(w_i \cdot u\right) \right)_{L^2(\Omega)},
\end{equation}
where $f \cdot g$ denotes pointwise multiplication, $f \ast g$ denotes convolution, and $\left(f, g\right)_{L^2(\Omega)}$ denotes the $L^2$ inner product on $\Omega$, for functions $f,g \in L^2(\Omega)$. Approximations of the form \eqref{eq:product_convolution} are known as \emph{product-convolution} approximations, because the action of each term in the sum consists of a pointwise product, followed by a convolution. To avoid nested subscripts, we write 
\begin{equation*}
	\convkernel_i := \convkernel_{x_i} \quad \text{and} \quad \impulseresponse_i := \impulseresponse_{x_i},
\end{equation*}
with $\convkernel_{x_i}$ defined in \eqref{eq:convolution_kernel} and $\impulseresponse_{x_i}$ defined in \eqref{eq:impulse_response_defn}, to denote local convolution kernels and impulse responses corresponding to a collection of points $\{x_i\}_{i=1}^\convrank \subset \Omega$ scattered throughout the domain. The functions $w_i$ are spatially varying weighting functions that are used to interpolate the convolution kernels $\convkernel_i$. Our operator approximation method is defined by how we compute the convolution kernels, $\convkernel_i$, how we choose the points, $x_i$, and what weighting functions, $w_i$, that we use. In Section \ref{sec:overview_intro} we summarize the answers to these questions; we will provide detailed answers to these questions in Section \ref{sec:method}.

@@

In order to choose the points $x_i$, we need to estimate the support of $\impulseresponse_i$ \emph{before} we compute it. The supports of the functions $\impulseresponse_x$ are estimated a-priori, for all $x \in \Omega$ simultaneously, via a procedure that involves applying $\Aop^T$ to a small number of specially chosen functions (three functions if $\gdim=1$, six if $\gdim=2$, ten if $\gdim=3$), then post processing the results (Section \ref{eq:mean_and_covariance_estimation}). This procedure is the only part of the paper that requires non-negativity of the integral kernel.

@@

Once the $\convkernel_i$ and $w_i$ are computed, we convert $\AopPc$ to hierarchical matrix (H-matrix) format. Let $\Aker_\text{pc}$ denote the integral kernel associated with $\AopPc$. While $\Aker(y,x)$ is not easily computable, $\Aker_\text{pc}(y,x)$ is given by the formula
\begin{equation}
	\label{eq:kernel_entries}
	\AkerPc(y,x) = \sum_{i=1}^\convrank \convkernel_i(y-x) w_i(x).
\end{equation}
This follows from~\eqref{eq:product_convolution} and the fact that the convolution operator $u \mapsto \convkernel_i \ast u$ has $(y,x)$ kernel entry given by $\convkernel(y-x)$.
Formula~\eqref{eq:kernel_entries} allows us to construct a H-matrix representation of $\AopPc$ using the conventional adaptive cross H-matrix construction method, in which one forms low-rank approximations of blocks of the matrix by sampling rows and columns of those blocks \cite{HACA}. Once in H-matrix format, fast H-matrix arithmetic is used to invert $\AopPc$, or perform other useful matrix operations.

@@

		\For{$k=1,2,\dots,\text{max\_batches}$}{
			Choose a batch of sample points, $\pointbatch_k$, using Algorithm \ref{alg:point_choice}

			Compute $\combresponse_k$ by applying $\Aop$ to the dirac comb for $\pointbatch_k$ (Section \ref{sec:get_impulse_response})

@@

The ellipsoid $E_x$ is not immediately available. To use $E_x$, we must first compute the scalar $\spatialvol(x)$, the length-$\gdim$ vector $\spatialmean(x)$, and the $\gdim \times \gdim$ matrix $\spatialcov(x)$ (recall $\gdim \in \{1,2,3\}$ is the spatial dimension of $\Omega$). The direct approach to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$. But this direct approach is computationally infeasible because and it requires one operator action of $\Aop$ per point $x$ that we wish to evaluate $E_x$ at, and we will need to evaluate $E_x$ at a large number of points. Fortunately, it is possible to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$, \emph{for all points $x$ simultaneously}, by applying $\Aop^T$ to one constant function, $\gdim$ linear functions, and $\gdim(\gdim+1)/2$ quadratic functions. We present this efficient method of computing $\spatialvol$, $\spatialmean$, and $\spatialcov$ in Theorem \ref{thm:vol_mean_cov} and summarize the method in Algorithm \ref{alg:varhpi_mean_cov}.

\begin{theorem}[Impulse response moments]
	\label{thm:vol_mean_cov}
	Let $\spatialvol:\Omega \rightarrow \mathbb{R}$, $\spatialmean:\Omega \rightarrow \mathbb{R}^\gdim$, and $\spatialcov:\Omega \rightarrow \mathbb{R}^{\gdim \times \gdim}$, be the following functions:
	\begin{equation*}
		\spatialvol(x) := \int_{\Omega} \impulseresponse_x(y) dy, \qquad
		\spatialmean(x) := \mathbb{E}(\widehat{\impulseresponse}_x), \qquad
		\spatialcov(x) := \Var(\widehat{\impulseresponse}_x),
	\end{equation*}
	where $\widehat{\impulseresponse}_x := \impulseresponse_x / \spatialvol(x)$ denotes the normalized version of $\impulseresponse_x$, $\mathbb{E}(\widehat{\impulseresponse}_x)$ denotes the mean of $\widehat{\impulseresponse}_x$, and $\Var(\widehat{\impulseresponse}_x)$ denotes the variance of $\widehat{\impulseresponse}_x$.	Let  $x^i$ denote the $i^\text{th}$ component of $x$, i.e., $x = \left(x^1, x^2, \dots, x^\gdim\right)$. Let $C$, $\{L^i\}_{i=1}^\gdim$, and ${\{Q^{ij}\}_{i=1}^\gdim}_{j=1}^\gdim$ be the following constant, linear, and quadratic functions:
	\begin{equation*}
		C(x) := 1, \qquad
		L^i(x) := x^i, \qquad
		Q^{ij}(x) = x^i x^j.
	\end{equation*}
	Let $f/g$ denote the pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ denote the pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$. The following results hold:
	\begin{subequations}
		\label{eq:vol_mean_var}
		\begin{align}
			\spatialvol =& \left(\Aop^T C\right)^* \\
			\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
			\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
		\end{align}
	\end{subequations}
	 for $i=1,\dots, \gdim$, $j=1,\dots,\gdim$.
\end{theorem}

@@

		\tcp{Compute mean $\spatialmean$}
		\For{$i=1,2,\dots,\gdim$}{
			Form linear function $L^i(x) = x^i$

			Compute $\spatialmean^i = \left(\Aop^T L^i\right)^* / \spatialvol$
		}
		\tcp{Compute covariance $\spatialcov$}
		\For{$i=1,2,\dots,\gdim$}{
			\For{$j=1,\dots,i$}{
				Form quadratic function $Q^{ij}(x) = x^i x^j$

@@

We use a greedy ellipsoid packing algorithm to choose batches of sample points, $\pointbatch_k$, such that there is no overlap between the ellipsoids, $E_{x_i}$, associated with the sample points, $x_i$, within a batch. These sample point batches will be used in Section \ref{sec:get_impulse_response} to compute many impulse responses of $\Aop$ using only a small number of operator actions of $\Aop$. The support of $\impulseresponse_x$ is (approximately) contained in the ellipsoid $E_x$, so by applying $\Aop$ to the Dirac comb associated with all sample points in a batch, we will recover the impulse responses for all sample points in that batch.

We start with a finite set of candidate points $P$. To build $\pointbatch_k$, first we initialize $\pointbatch_k$ as an empty set. Then we select a candidate point $p \in P$ that is the farthest away from all points in previous sample point batches. That is, $p$ is a maximizer of the following optimization problem:
\begin{equation*}
\max_{p \in P} \min_{x \in \pointbatch_1 \cup \dots \cup \pointbatch_{k-1}} \|p - x\|.
\end{equation*}
Candidate points for the first batch are chosen arbitrarily from $P$.
Once $p$ is selected, we remove $p$ from $P$. If $p$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_p \cap E_q = \{\}$ for all $q \in \pointbatch_k$, then we add $p$ to $\pointbatch_k$. Otherwise we discard $p$. This process repeats until there are no more points in $P$.  This is detailed in Algorithm \ref{alg:point_choice}. We determine whether $E_p \cap E_q = \{\}$ using the fast ellipsoid intersection test described in Appendix \ref{sec:fast_ellipsoid_intersection_test}.

We repeat the process to construct several batches of points $\pointbatch_1, \pointbatch_2, \dots$, until the number of batches exceeds a desired threshold. In our implementation, for each batch the set of candidate points $P$ is initialized as the set of all Lagrange nodes for the finite element basis functions used to discretize the problem, except those points in previously chosen batches.

@@

	\Input{Finite set of candidate points $P\subset \Omega$, \\spatialmean $\spatialmean(x)$ and covariance $\spatialcov(x)$, \\domain boundary information $\partial \Omega$, \\previous sample point batches $\pointbatch_1, \dots, \pointbatch_{k-1}$}
	\Output{Batch of new sample points $\pointbatch_k$}

	Initialize empty new batch of sample points, $\pointbatch_k = \{\}$

	\While{$P$ is not empty}{
		Determine the point $p \in P$ that is farthest from all points in the previous sample point batches $\pointbatch_1,\dots,\pointbatch_{k-1}$

@@

		\If{$E_p \cap E_q \neq \{\}$ for all $q \in \pointbatch_k$ and $E_p \cap \partial \Omega = \{\}$}{
			\tcp{$E_p$ and $E_q$ are the ellipsoids defined in \eqref{eq:support_ellipsoid}}

			Add $p$ to $\pointbatch_k$

@@

    \SetKwFunction{FMain}{point\_is\_acceptable}
	\caption{Choosing one batch of sample points, $\pointbatch_k$}
	\label{alg:point_choice}
\end{algorithm2e}

@@

We compute convolution kernels $\convkernel_{x_i}$ in batches by applying $\Aop$ to a small number Dirac combs. Each application of $\Aop$ to a Dirac comb yields one batch of impulse responses, $\impulseresponse_{x_i}$, which yields one batch of convolution kernels, $\convkernel_{x_i}$. The Dirac comb, $\diraccomb_k$, associated with a batch of sample points, $\pointbatch_k$, is the sum of Dirac distributions (point sources) centered at the points $x_i \in \pointbatch_k$. That is,
\begin{equation*}
	\diraccomb_k := \sum_{x_i \in \pointbatch_k} \delta_{x_i}.
\end{equation*}
For each batch $\pointbatch_k$, we compute the action of $\Aop$ on the associated Dirac comb:
\begin{equation}
	\label{eq:dirac_comb_H_action}
	\combresponse_k := \left(\Aop \diraccomb_k^*\right)^*.
\end{equation}
By linearity, the resulting function $\combresponse_k$ may be written as
\begin{equation}
	\label{eq:phi_b}
	\combresponse_k = \left(\Aop \left(\sum_{x_i \in \pointbatch_k} \delta_{x_i}\right)^*\right)^* = \sum_{x_i \in \pointbatch_k} \left(\Aop \delta_{x_i}^*\right)^* = \sum_{x_i \in \pointbatch_k} \impulseresponse_{x_i}.
\end{equation}
Since the points $x_i$ are chosen so that the ellipsoid $E_{x_i}$ that (approximately) supports $\impulseresponse_i$, and the ellipsoid $E_{x_j}$ that (approximately) supports $\impulseresponse_j$ do not overlap when $i \neq j$, from \eqref{eq:convolution_kernel} and \eqref{eq:phi_b} we have (approximately)
\begin{equation}
\label{eq:varphi_eval}
	\convkernel_{x_i}(z) = \impulseresponse_i(z+x_i) = 
	\begin{cases}
		\combresponse_k(z+x_i), & z+x_i \in E_{x_i} \\
		0, & \text{otherwise}
	\end{cases}
\end{equation}
for all $x_i \in \pointbatch_k$. By performing one matrix-vector product, \eqref{eq:dirac_comb_H_action}, we recover the convolution kernels $\convkernel_{x_i}$ associated with every point $x_i \in \pointbatch_k$.

@@

To convert $\AopPc$ to H-matrix format (Section \ref{sec:H_matrix_conversion}), we will need to evaluate $\convkernel_{x_i}(z)$ for a large number of points $z$. Per \eqref{eq:varphi_eval}, evaluating $\convkernel_{x_i}(z)$ requires evaluating $\combresponse_k(z+x_i)$. In computations, $\combresponse_k$ will be a finite element function formed on an irregular mesh, and $z+x_i$ will typically not be a gridpoint of the mesh. Evaluating $\combresponse_k(z+x_i)$ therefore requires finding which mesh cell that $z+x_i$ is in, then evaluating several polynomial functions to evaluate the finite element function on that cell. Although this procedure is asymptotically scalable, it is slow in practice \cite{FEMEVALSLOW}.

To speed up evaluation of $\convkernel_{x_i}$, we interpolate $\convkernel_{x_i}$ onto a regular rectilinear grid on a coordinate axis aligned box that just barely contains the ellipsoid $E_{x_i}$. We transfer the function $\convkernel_{x_i}$ from the irregular mesh to the regular grid by evaluating $\convkernel_{x_i}$ at all gridpoints in the regular grid, using the slow process described in the previous paragraph. This is an upfront cost which is done ``offline.'' After $\convkernel_{x_i}$ has been transferred to the regular grid, we use fast regular grid interpolation to perform all subsequent ``online'' evaluations of $\convkernel_{x_i}$. This yields a substantial overall speedup because the number of finite element function evaluations required to transfer the functions $\convkernel_{x_i}$ to the regular grid is $O(\fedim)$, while the number of kernel evaluations required to convert $\AopPc$ to H-matrix format is $O(k \fedim \log \fedim)$. Here $\fedim$ is the number of finite element degrees of freedom, and $k$ is the H-matrix rank. Furthermore, transferring $\convkernel_{x_i}$ to a regular grid requires evaluating $\convkernel_{x_i}$ at a predetermined and highly structured collection of points. This structure allows for faster determination of which mesh cells that points are in, and allows one to evaluate the finite element function at all regular gridpoints in parallel. In contrast, constructing the H-matrix requires evaluating $\convkernel_{x_i}$ at an unpredictable sequence of scattered of points, which is an expensive task for finite element function evaluation, but a cheap task for regular grid interpolation. 
Having $\convkernel_{x_i}$ stored on a regular grid also simplifies the implementation of the method that we use to deal with boundary artifacts, which will be described in Section \ref{sec:boundary_extension}.

@@

Consider the following optimization problem,
\begin{equation}
	\label{eq:wi_optimization_problem_generic}
	\begin{aligned}
	\min_{v \in H^2_N(\Omega)} &\quad \frac{1}{2}\norm{\Delta v}_{L^2(\Omega)}^2 \\
	\text{such that} &\quad v(x_j) = \interpolatedvalues_j, \quad j=1,\dots,\convrank,
	\end{aligned}
\end{equation}
in which one seeks a smooth function $v$ that takes the values $\interpolatedvalues_j \in \mathbb{R}$ at the points $x_j$. Here $\Delta$ is the Laplacian on $\Omega$ and $H^2_N(\Omega)$ is the space of all functions in $H^2(\Omega)$ with Neumann zero boundary conditions. We choose the $i^\text{th}$ weighting function, $w_i(x)$, to be the solution to optimization problem \eqref{eq:wi_optimization_problem_generic}, with 
\begin{equation*}
	b_j = w_i(x_j) = 
	\begin{cases}
		1, & j=i\\
		0, & j \neq i.
	\end{cases}
\end{equation*}
Using these weighting functions yields the ``flattest possible'' interpolation of the convolution kernels $\convkernel_i$, in a least-squares sense (Theorem \ref{thm:smoothest_interpolant}). Also, these weighting functions sum to one pointwise (Proposition \ref{thm:wi_sum_to_one}).

%For each $i=1,\dots,\convrank$, optimization problem \eqref{eq:wi_optimization_problem} is an example of the following generic optimization problem:
%\begin{equation}
%	\label{eq:wi_optimization_problem_generic}
%	\begin{aligned}
%		\min_{v \in H^2_N(\Omega)} &\quad \frac{1}{2}\norm{\Delta v}_{L^2(\Omega)}^2 \\
%		\text{such that} &\quad v(x_j) = \interpolatedvalues_j, \quad j=1,\dots,\convrank,
%	\end{aligned}
%\end{equation}
%where $\interpolatedvalues_j \in \mathbb{R}$, $j=1,\dots,\convrank$ are given function values that $v$ must take at the points $x_j$.
